Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-20 manuscript/manuscript.Rmd
install 7zip

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  texlive: true
  sysdeps: 
    - apt-utils
    - grass
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

# Optional: Acknowledgements
acknowledgements: |
  This work was supported by the U.S. National Science Foundation’s Dynamics of Coupled Natural and Human Systems (CNH) e Program (award 1517823) and Macrosystems Biology Program (awards EF-1638679, EF-1638554, EF-1638539, and EF-1638550). PAS was also supported by the USDA National Institute of Food and Agriculture, Hatch project 1013544. We thank the following individuals for a friendly review of an earlier draft: K.S. Cheruvelil, I.M. McCullough, N. Smith, and A. Poisson.

@@

Lake phosphorus (P) retention is an important characteristic of lakes that can be used to predict P concentrations and to evaluate lake sensitivity to nutrient O loading and eutrophication (Alexander et al. 2008; Milstead et al. 2013). Specifically, P retention is an integrated measure of internal P losses including permanent sedimentation, biological uptake, and other processes that remove P from the water column \citep{chapra2008surface}. P retention has been well studied in lakes because P determines lake trophic status and downstream watershed yields (i.e. export to terminal lakes and coastal estuaries). Although previous studies have shown that lake P retention is related to water residence time (Vollenweider 1975), large uncertainties exist around this relationship (Brett and Benjamin 2007, Milstead et al. 2013). These uncertainties can be large in lakes with intermediate water residence times, particularly compared to lakes with either extremely short or extremely long water residence times (Figure A1). For example, lakes with very long water residence times (on the order of a decade or longer) have complete or near complete P retention, while lakes with very short water residence times (on the order of days) have almost no P retention (Brett and Benjamin, 2007). The reason for the substantial uncertainty in lake P retention between these two extremes may be that predicting retention solely on the basis of water residence time does not capture many of the other factors and processes that affect P retention (Figure \ref{fig1}).

@@

One well-studied factor that has been shown to influence lake P retention is lake depth through its influence on internal processing of P loads (Søndergaard et al. 2013; Vollenweider 1975). The mechanism for such an influence is that lake depth controls thermal stratification and material resuspension from the benthos. As a result, shallow lakes have a tendency to mix throughout the summer causing redistribution e of sedimented phosphorus throughout the mixed zone (Fee et al. 1996). This mixing and redistribution of sedimented P often leads to tighter benthic-pelagic coupling and increased P recycling e (Cha et al. 2013). Thus, depth is one example of a lake characteristic that is likely to influence P retention in concert with water residence time (Cheng et al. 2010).

@@

In general, point source inputs e to lakes are associated with increased dissolved P loads (Kronvang 1992; Russell et al. 1998), whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land e use cover, typically have higher particulate P loads (Carpenter et al. 1998; Sharpley et al. 1994). Exceptions to these generalizations have been observed in areas where increased dissolved P loading O occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release (Bennett and Carpenter 2001; Powers et al. 2016). Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics (Jarvie et al. 2011). Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints (Guy et al. 1994). As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett 2006, Djodjic and Markensten 2018).

Apart from land-use cover, another potential proxy for particulate load fraction entering lakes is the type and amount of freshwater connections to a lake, which we argue is easier to measure than other proxies, and could help to improve estimates of lake P retention, especially in lakes for which we lack P loading data (Figure 1). Although freshwater connectivity may be easy to measure from a logistical standpoint, there are still many ways to measure connectivity that likely represent different mechanisms of water and material flow (Figure 2). We broadly define and study two types of freshwater connectivity that correspond to either stream-based metrics or lake-stream-based metrics. First, lake-stream-based metrics measure the connections between a lake, other upstream lakes, and streams e in their watershed. This type of connectivity can be quantified by measuring the closest distance to an upstream stream-connected lake (Figure 2A), or by measuring the total upstream lake area e (Figure 2B). Second, stream-based metrics measure the connections between inflowing streams and their surrounding land whereby increasing stream connections lead to a greater abundance of land-water O interfaces and greater transport of material from hillslope and riparian areas adjacent to watershed stream networks (Figure 2C-E).

@@

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network (Covino 2017). For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance (Gomi et al. 2002). Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. e For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes e (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

We asked two questions in this  study: 1) Which measures of freshwater connectivity influence lake phosphorus retention? 2) What e spatial extent of connectivity most strongly influences P retention? To answer these questions, we fit statistical P retention models in a Bayesian hierarchical framework following Cheng et O al. (2010) where 2 separate values of the k processing parameter were estimated for lakes with either high or low values of each connectivity metric. Using this approach, higher k values for a specific lake connectivity class indicates more extensive in-lake processing and higher P retention. We applied this model to a dataset of 129 lakes across a wide range of hydrologic, geologic, and climatic settings. We fit separate models using each combination of connectivity metric and spatial extent in an effort to determine whether P retention is more strongly controlled at the lake subwatershed extent or the lake watershed extent.

@@

We used data on P retention, maximum depth, and water residence time from 129 lakes in the National Eutrophication Survey (USEPA 1975, Stachelek et al. 2018). Mean annual P loading, P discharge, and P retention values in the National Eutrophication Survey (NES) dataset were calculated based on monthly sampling for P in tributary and outlet discharge points as well as any municipal waste discharges from 1972 to 1975. Here, P retention is a unitless value representing the fraction of incoming P loads. Sampling frequency for water discharge and residence time varied among lakes but details of these variations were not provided in the source dataset (USEPA 1975, Stachelek et al. 2018). Estimates of water discharge and residence time in the NES dataset represent normalized mean flow estimates expected to occur during a period of average precipitation and hydrology (USEPA 1975). Water residence time for our study lakes ranged from 1 week to 17 years with an interquartile e range of 3 months to 1.8 years while P retention ranged from 0.06 to 0.99 with an interquartile range of 0.24 to 0.59 (Table 1).

We supplemented the NES dataset with boundaries O for lake subwatersheds, as well as estimates of stream density, upstream lake area, upstream lake connection(s), baseflow (an index of groundwater inputs), land-use cover, and other water quality measurements from the LAGOS-NE dataset (Table 1; Soranno et al. 2017). Our study lakes encompassed a wide range of land-use cover types and nutrient levels (Table 1). Although, lake subwatersheds were variable with respect to agricultural land use cover, we did not observe a strong relationship with lake P retention (Figure A2). On average, the water quality (total phosphorus, chlorophyll concentration, and Secchi depth) of the lakes in our study are similar to other US lakes as measured by the stratified random sampling design of the National Lakes Assessment (NLA) lake population (USEPA 2016). However, our lakes are substantially larger and deeper than most NLA lakes (Figure A3).

@@

In addition to data from the NES and  LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related e to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing  aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo 2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we
expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal
to 10 ha (0.1 km2). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input (USGS e 2018). Average link length was calculated as the total stream length in a given watershed divided v
 by the number of stream reaches after dissolving (removing) any network points that do not e occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes)
expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams (Barbera and Rosso 1989). Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively (Stachelek 2018a, Stachelek 2018b). The algorithms in the `streamnet` package use the `sf` R package (Pebesma 2018) as well as the `v.net` and `v.stream.order` modules (Jasiewicz and Metz 2011) included in GRASS GIS (GRASS Development Team 2017). All processed connectivity data and code are available at [Zenodo DOI].

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each connectivity metric based on the bivariate e relationship between each metric and P retention. Prior studies have used a similar binary splitting approach to examine the effect of various exogenous factors on lake P retention (Cheng et al. 2010; Shimoda O and Arhonditsis 2015). We determined splitting criteria for each metric from the results of a random forest procedure incorporating conditional inference trees (ctree, Hothorn and Zeileis 2015). This procedure creates binary splits of the independent variables (i.e. each of the connectivity metrics), which are recursively repeated to find the split that maximizes association with the dependent variable (P retention). The advantage of the ctree technique over the more typical classification-regression tree (CART) technique is that tree growth stopping rules are pre-specified (Hothorn et al., 2006). As a result, some of subjectivity associated with post-hoc tree pruning is avoided.

@@

We examined the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention to determine the potential for strong relationships among any of the variables which were not accounted for by our model structure (Eq 2, Figure 4). We found evidence for some relationships among these variables, but none that suggest either redundancy among connectivity metrics or that otherwise limit our ability to infer relationships with P retention (Figure 4). For example, the Pearson correlation (r) between water residence time and stream density, which is implicitly accounted for in our model structure, was 0.29 (p < 0.05). In contrast, the correlation between water residence time and upstream lake area, which is unaccounted for in our model structure, e was only 0.10 (p > 0.05). Note that interactions between connectivity metrics and water residence time are accounted for in our hierarchical model structure because Eq. 2 uses a global coefficient e x for water residence time that is estimated separately from the hierarchical and connectivity-dependent P decay coefficient k. This is conceptually similar to fitting the k-connectivity-P retention O relationship to the residuals of the water residence time to P retention relationship. Perhaps surprisingly, we did not observe strong correlations between P loading and water residence time (r = 0.13, p > 0.05) or between lake depth and most connectivity metrics (r < 0.17). This suggests that our P retention results are not confounded by variations in lake depth, by interactions between connectivity and P loading, or by interactions between water residence time and P loading.

@@

A secondary purpose of examining the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention was to determine if our connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among connectivity metrics were low and of a similar magnitude as the correlations between connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation between any connectivity metric or lake characteristic was upstream lake area and lake watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake characteristics not accounted for in Eq 2. These include the correlation between P loading and lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake subwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and upstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P loading, watershed area, and upstream e lake area, we did not observe a strong correlation between P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time e and P retention (r = 0.44, p > 0.05). Our observation that the correlation between P retention and P loading was not appreciably stronger
than correlations between P retention and our connectivity O metrics suggests that our estimates of connectivity metric effects on P retention are not confounded by interactions between water residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found that lakes with similar connectivity metric values, were spatially clustered (Figure 5). In particular, we found that lakes were concentrated in either the southern or northern portions of our study area depending on their connectivity metric value (Figure 5). This observation is consistent with the findings of Fergus et al. (2017) that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our study area.

@@

We found that some connectivity metrics were more strongly related to P retention than others. For example, the model $R^{2}$ for network e average link length was higher ($R^{2}$ = 0.41) than the global model ($R^{2}$ = 0.34). For other connectivity metrics, such as network stream order ratio,
goodness-of-fit ($R^{2}$ = 0.36) was very similar to the global O model ($R^{2}$ = 0.34). Model fit for other connectivity metrics was in between these two extremes (0. 36 < $R^{2}$ < 0.41). Overall, we found that all hierarchical models had at least a marginally better fit to the water residence time versus P retention relationship than a global model which does not account for connectivity (Figure 6).
Although we found a discernable effect of connectivity metrics on lake P retention, the somewhat modest improvements in model fit may be due to the fact that water residence time remains a dominant effect even after accounting for freshwater connectivity.

@@

We found that connectivity metrics e measured at the lake watershed extent were more strongly associated with P retention than metrics measured at the lake subwatershed extent (Table 2, Figure 7). Specifically, the metrics e that had the strongest association with P retention such as average link length ($\Delta$k = 0.23), closest lake distance ($\Delta$k = 0.22), and stream density ($\Delta$k = 0.20) also had a stronger association at the lake watershed O extent rather than at the lake subwatershed extent (Table 2, Figure 7). An exception to this pattern was observed with the baseflow connectivity metric where although greater differences at the lake subwatershed extent
were more strongly associated with P retention, the sign of the effect was variable depending on measurement extent (i.e. the value of the P decay parameter was positively related to connectivity metric values at the subwatershed extent but was negatively related at the watershed extent). For several connectivity metrics, we judged that differences in P decay among lakes with either low or high connectivity metric values were not significant because they translated to differences in P retention that were less than measurement precision (Table 2).

@@

Although prior studies have found that lake P retention is related to water residence time (Brett and Benjamin, Cheng, et al. 2010, Vollenweider 1975), there is substantial variation around this relationship, particularly at intermediate water residence times. We found that some of this remaining variation could be explained using a hierarchical modelling framework that accounts for differences in freshwater connectivity among lakes. Although we found that the magnitude of this effect depends on the specific connectivity metric, our results are consistent with the findings of previous studies  showing that connectivity metrics are associated with differences in lake carbon input fluxes and differences in lake nitrogen output fluxes (Cardille et al. 2007; Schmadel et al. 2018). We also found e important differences in the association between connectivity metrics and P retention at different measurement extents. Specifically, we found that P retention was more strongly associated e with connectivity measured at the broader lake watershed extent rather than connectivity measured at the finer lake subwatershed extent.

>>>>

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  texlive: true
  sysdeps: 
    - grass
    - p7zip-full
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

# Optional: Acknowledgements
acknowledgements: |
  This work was supported by the U.S. National Science Foundation’s Dynamics of Coupled Natural and Human Systems (CNH) Program (award 1517823) and Macrosystems Biology Program (awards EF-1638679, EF-1638554, EF-1638539, and EF-1638550). PAS was also supported by the USDA National Institute of Food and Agriculture, Hatch project 1013544. We thank the following individuals for a friendly review of an earlier draft: K.S. Cheruvelil, I.M. McCullough, N. Smith, and A. Poisson.

@@

<!-- \bibliography{pinp} --->

Lake phosphorus (P) retention is an important characteristic of lakes that can be used to predict P concentrations and to evaluate lake sensitivity to nutrient loading and eutrophication (Alexander et al. 2008; Milstead et al. 2013). Specifically, P retention is an integrated measure of internal P losses including permanent sedimentation, biological uptake, and other processes that remove P from the water column \citep{chapra2008surface}. P retention has been well studied in lakes because P determines lake trophic status and downstream watershed yields (i.e. export to terminal lakes and coastal estuaries). Although previous studies have shown that lake P retention is related to water residence time \citep{vollenweiderInputoutputModels1975}, large uncertainties exist around this relationship (Brett and Benjamin 2007, Milstead et al. 2013). These uncertainties can be large in lakes with intermediate water residence times, particularly compared to lakes with either extremely short or extremely long water residence times (Figure A1). For example, lakes with very long water residence times (on the order of a decade or longer) have complete or near complete P retention, while lakes with very short water residence times (on the order of days) have almost no P retention (Brett and Benjamin, 2007). The reason for the substantial uncertainty in lake P retention between these two extremes may be that predicting retention solely on the basis of water residence time does not capture many of the other factors and processes that affect P retention (Figure \ref{fig1}).

@@

One well-studied factor that has been shown to influence lake P retention is lake depth through its influence on internal processing of P loads (Søndergaard et al. 2013; Vollenweider 1975). The mechanism for such an influence is that lake depth controls thermal stratification and material resuspension from the benthos. As a result, shallow lakes have a tendency to mix throughout the summer causing redistribution of sedimented phosphorus throughout the mixed zone (Fee et al. 1996). This mixing and redistribution of sedimented P often leads to tighter benthic-pelagic coupling and increased P recycling (Cha et al. 2013). Thus, depth is one example of a lake characteristic that is likely to influence P retention in concert with water residence time (Cheng et al. 2010).

@@

In general, point source inputs to lakes are associated with increased dissolved P loads (Kronvang 1992; Russell et al. 1998), whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land use cover, typically have higher particulate P loads (Carpenter et al. 1998; Sharpley et al. 1994). Exceptions to these generalizations have been observed in areas where increased dissolved P loading occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release (Bennett and Carpenter 2001; Powers et al. 2016). Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics (Jarvie et al. 2011). Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints (Guy et al. 1994). As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett 2006, Djodjic and Markensten 2018).

Apart from land-use cover, another potential proxy for particulate load fraction entering lakes is the type and amount of freshwater connections to a lake, which we argue is easier to measure than other proxies, and could help to improve estimates of lake P retention, especially in lakes for which we lack P loading data (Figure 1). Although freshwater connectivity may be easy to measure from a logistical standpoint, there are still many ways to measure connectivity that likely represent different mechanisms of water and material flow (Figure 2). We broadly define and study two types of freshwater connectivity that correspond to either stream-based metrics or lake-stream-based metrics. First, lake-stream-based metrics measure the connections between a lake, other upstream lakes, and streams in their watershed. This type of connectivity can be quantified by measuring the closest distance to an upstream stream-connected lake (Figure 2A), or by measuring the total upstream lake area (Figure 2B). Second, stream-based metrics measure the connections between inflowing streams and their surrounding land whereby increasing stream connections lead to a greater abundance of land-water interfaces and greater transport of material from hillslope and riparian areas adjacent to watershed stream networks (Figure 2C-E).

@@

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network (Covino 2017). For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance (Gomi et al. 2002). Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

We asked two questions in this study: 1) Which measures of freshwater connectivity influence lake phosphorus retention? 2) What spatial extent of connectivity most strongly influences P retention? To answer these questions, we fit statistical P retention models in a Bayesian hierarchical framework following Cheng et al. (2010) where 2 separate values of the k processing parameter were estimated for lakes with either high or low values of each connectivity metric. Using this approach, higher k values for a specific lake connectivity class indicates more extensive in-lake processing and higher P retention. We applied this model to a dataset of 129 lakes across a wide range of hydrologic, geologic, and climatic settings. We fit separate models using each combination of connectivity metric and spatial extent in an effort to determine whether P retention is more strongly controlled at the lake subwatershed extent or the lake watershed extent.

@@

We used data on P retention, maximum depth, and water residence time from 129 lakes in the National Eutrophication Survey (USEPA 1975, Stachelek et al. 2018). Mean annual P loading, P discharge, and P retention values in the National Eutrophication Survey (NES) dataset were calculated based on monthly sampling for P in tributary and outlet discharge points as well as any municipal waste discharges from 1972 to 1975. Here, P retention is a unitless value representing the fraction of incoming P loads. Sampling frequency for water discharge and residence time varied among lakes but details of these variations were not provided in the source dataset (USEPA 1975, Stachelek et al. 2018). Estimates of water discharge and residence time in the NES dataset represent normalized mean flow estimates expected to occur during a period of average precipitation and hydrology (USEPA 1975). Water residence time for our study lakes ranged from 1 week to 17 years with an interquartile range of 3 months to 1.8 years while P retention ranged from 0.06 to 0.99 with an interquartile range of 0.24 to 0.59 (Table 1).

We supplemented the NES dataset with boundaries for lake subwatersheds, as well as estimates of stream density, upstream lake area, upstream lake connection(s), baseflow (an index of groundwater inputs), land-use cover, and other water quality measurements from the LAGOS-NE dataset (Table 1; Soranno et al. 2017). Our study lakes encompassed a wide range of land-use cover types and nutrient levels (Table 1). Although, lake subwatersheds were variable with respect to agricultural land use cover, we did not observe a strong relationship with lake P retention (Figure A2). On average, the water quality (total phosphorus, chlorophyll concentration, and Secchi depth) of the lakes in our study are similar to other US lakes as measured by the stratified random sampling design of the National Lakes Assessment (NLA) lake population (USEPA 2016). However, our lakes are substantially larger and deeper than most NLA lakes (Figure A3).

@@

In addition to data from the NES and LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo 2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we
expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal
to 10 ha (0.1 km2). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input (USGS 2018). Average link length was calculated as the total stream length in a given watershed divided v
 by the number of stream reaches after dissolving (removing) any network points that do not occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes)
expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams (Barbera and Rosso 1989). Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively (Stachelek 2018a, Stachelek 2018b). The algorithms in the `streamnet` package use the `sf` R package (Pebesma 2018) as well as the `v.net` and `v.stream.order` modules (Jasiewicz and Metz 2011) included in GRASS GIS (GRASS Development Team 2017). All processed connectivity data and code are available at [Zenodo DOI].

@@

Our lake connectivity classes were formed by dividing the lake dataset into two classes for each connectivity metric based on the bivariate relationship between each metric and P retention. Prior studies have used a similar binary splitting approach to examine the effect of various exogenous factors on lake P retention (Cheng et al. 2010; Shimoda and Arhonditsis 2015). We determined splitting criteria for each metric from the results of a random forest procedure incorporating conditional inference trees (ctree, Hothorn and Zeileis 2015). This procedure creates binary splits of the independent variables (i.e. each of the connectivity metrics), which are recursively repeated to find the split that maximizes association with the dependent variable (P retention). The advantage of the ctree technique over the more typical classification-regression tree (CART) technique is that tree growth stopping rules are pre-specified (Hothorn et al., 2006). As a result, some of subjectivity associated with post-hoc tree pruning is avoided.

@@

We examined the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention to determine the potential for strong relationships among any of the variables which were not accounted for by our model structure (Eq 2, Figure 4). We found evidence for some relationships among these variables, but none that suggest either redundancy among connectivity metrics or that otherwise limit our ability to infer relationships with P retention (Figure 4). For example, the Pearson correlation (r) between water residence time and stream density, which is implicitly accounted for in our model structure, was 0.29 (p < 0.05). In contrast, the correlation between water residence time and upstream lake area, which is unaccounted for in our model structure, was only 0.10 (p > 0.05). Note that interactions between connectivity metrics and water residence time are accounted for in our hierarchical model structure because Eq. 2 uses a global coefficient x for water residence time that is estimated separately from the hierarchical and connectivity-dependent P decay coefficient k. This is conceptually similar to fitting the k-connectivity-P retention relationship to the residuals of the water residence time to P retention relationship. Perhaps surprisingly, we did not observe strong correlations between P loading and water residence time (r = 0.13, p > 0.05) or between lake depth and most connectivity metrics (r < 0.17). This suggests that our P retention results are not confounded by variations in lake depth, by interactions between connectivity and P loading, or by interactions between water residence time and P loading.

@@

A secondary purpose of examining the bivariate relationships between connectivity, water residence time, P loading and other factors related to P retention was to determine if our connectivity metrics were related to each other such that they provide similar information. With the exception of stream density and baseflow (r = -0.52, p < 0.05), correlations among connectivity metrics were low and of a similar magnitude as the correlations between connectivity metrics and water residence time (0.10 < r < 0.29). The strongest correlation between any connectivity metric or lake characteristic was upstream lake area and lake watershed area (r = 0.77, p < 0.05).
We observed notably strong correlations between P loading and several lake characteristics not accounted for in Eq 2. These include the correlation between P loading and lake watershed area (r = 0.62, p < 0.05) as well as the correlation between P loading and lake subwatershed area (r = o
 0.76, p < 0.05). In addition, the correlation between P loading and upstream lake area was quite strong (r = 0.52, p < 0.05). Despite strong correlations between P loading, watershed area, and upstream lake area, we did not observe a strong correlation between P loading and P retention (r = 0.15, p > 0.05). Notably, this correlation was much weaker than the correlation between water residence time and P retention (r = 0.44, p > 0.05). Our observation that the correlation between P retention and P loading was not appreciably stronger
than correlations between P retention and our connectivity metrics suggests that our estimates of connectivity metric effects on P retention are not confounded by interactions between water residence time and P loading.
Although we did not observe strong correlations among connectivity metrics, we found that lakes with similar connectivity metric values, were spatially clustered (Figure 5). In particular, we found that lakes were concentrated in either the southern or northern portions of our study area depending on their connectivity metric value (Figure 5). This observation is consistent with the findings of Fergus et al. (2017) that lakes in the northern portion of our study area have distinct freshwater connectivity as compared to lakes in the southern portion of our study area.

@@

We found that some connectivity metrics were more strongly related to P retention than others. For example, the model $R^{2}$ for network average link length was higher ($R^{2}$ = 0.41) than the global model ($R^{2}$ = 0.34). For other connectivity metrics, such as network stream order ratio,
goodness-of-fit ($R^{2}$ = 0.36) was very similar to the global model ($R^{2}$ = 0.34). Model fit for other connectivity metrics was in between these two extremes (0. 36 < $R^{2}$ < 0.41). Overall, we found that all hierarchical models had at least a marginally better fit to the water residence time versus P retention relationship than a global model which does not account for connectivity (Figure 6).
Although we found a discernable effect of connectivity metrics on lake P retention, the somewhat modest improvements in model fit may be due to the fact that water residence time remains a dominant effect even after accounting for freshwater connectivity.

@@

We found that connectivity metrics measured at the lake watershed extent were more strongly associated with P retention than metrics measured at the lake subwatershed extent (Table 2, Figure 7). Specifically, the metrics that had the strongest association with P retention such as average link length ($\Delta$k = 0.23), closest lake distance ($\Delta$k = 0.22), and stream density ($\Delta$k = 0.20) also had a stronger association at the lake watershed extent rather than at the lake subwatershed extent (Table 2, Figure 7). An exception to this pattern was observed with the baseflow connectivity metric where although greater differences at the lake subwatershed extent
were more strongly associated with P retention, the sign of the effect was variable depending on measurement extent (i.e. the value of the P decay parameter was positively related to connectivity metric values at the subwatershed extent but was negatively related at the watershed extent). For several connectivity metrics, we judged that differences in P decay among lakes with either low or high connectivity metric values were not significant because they translated to differences in P retention that were less than measurement precision (Table 2).

@@

Although prior studies have found that lake P retention is related to water residence time (Brett and Benjamin, Cheng, et al. 2010, Vollenweider 1975), there is substantial variation around this relationship, particularly at intermediate water residence times. We found that some of this remaining variation could be explained using a hierarchical modelling framework that accounts for differences in freshwater connectivity among lakes. Although we found that the magnitude of this effect depends on the specific connectivity metric, our results are consistent with the findings of previous studies showing that connectivity metrics are associated with differences in lake carbon input fluxes and differences in lake nitrogen output fluxes (Cardille et al. 2007; Schmadel et al. 2018). We also found important differences in the association between connectivity metrics and P retention at different measurement extents. Specifically, we found that P retention was more strongly associated with connectivity measured at the broader lake watershed extent rather than connectivity measured at the finer lake subwatershed extent.
